Mechanisation of Session-based Lambda Calculus Type Systems

### 2014-12-11 paper/paper.tex
Back to regular append; comma too special. Added paper beginnings.

\documentclass{mprop}

% Include Coq lstlisting language definition and things upon which it depends.
\usepackage{macros}
\usepackage{hyperref}

\usepackage{listings}
\usepackage{lstcoq}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{cmll}
\usepackage{prooftree}
\usepackage{amsmath}
\newcommand{\typestate}[2]{{#1}\gg{#2}}
\usepackage[style=alphabetic,sorting=nymt,natbib=true,
  maxbibnames=100]{biblatex}

\usepackage{tabularx}
\usepackage{tabulary}

@@

\newcommand{\Deliverable}[2]
           {\multicolumn{2}{|c|}{\textbf{D{#1}: {#2}}} \\ \hline}
\newcommand{\DSubD}[3]{D{#1}.{#2} & #3 \\ \hline}
\newcommand{\DTask}[6]{D#1T#2 & #3 & #4 & #5 & #6 \\ \hline}
\newcommand{\IDTsk}[2]{Task D{#1}T{#2}}

% Thanks to Stefan Kottwitz on tex.stackexhange.com:
% http://tex.stackexchange.com/questions/69662/how-to-globally-change-the-spacing-around-equations
%% \expandafter\def\expandafter\normalsize\expandafter{%
%%     \normalsize
%%     \setlength\abovedisplayskip{0pt}
%%     \setlength\belowdisplayskip{0pt}
%%     \setlength\abovedisplayshortskip{0pt}
%%     \setlength\belowdisplayshortskip{0pt}
%% }

\addbibresource{proposal.bib}

@@

\allowdisplaybreaks

% alternative font if you prefer
%\usepackage{times}

% for alternative page numbering use the following package
% and see documentation for commands
%\usepackage{fancyheadings}

@@

% other potentially useful packages
%\uspackage{amssymb,amsmath}
%\usepackage{url}
%\usepackage{fancyvrb}
%\usepackage[final]{pdfpages}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Mechanisation of Session-based Lambda Calculus Type Systems}
\author{Craig McLaughlin}
\date{\today}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\tableofcontents
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{intro}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Statement of Problem}

The aim of this project is to develop a framework for formalising lambda
calculus type systems involving session types using interactive theorem
proving (in particular, the system Coq). The motivation for the framework is
that it will enable a whole class of related languages and type systems to be
formalised with little extra proof effort. Inspired by previous work on a
mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}, I
shift to the lambda calculus to provide a basis on which session types for
mainstream programming languages can be studied, whilst taking advantage of
previous proof effort and specialising only for those parts unique to the
particular system. The framework will be heavily based on calculi developed by
\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
classical linear logic (CP) and a functional language with session types
(GV). GV is a minimal functional language supporting linearity so I believe
such a basis will be general enough to support reasoning about a large class
of functional languages with similar base properties.

From this framework, a number of directions can be explored including session
polymorphism session subtyping, and permitting different forms of aliasing
within the type system. The recent work by \citeauthor{Lindley:2014:SAP}
\cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding,
replication and polymorphism, would be an interesting place to start when
looking to add these extensions.

Additionally, the majority of previous session type systems for mainstream
programming languages have focussed exclusively on linear type systems. One
disadvantage of a linear type system is the need to re-bind objects after each
operation~\cite{Gay:2010:LAST}. One could extend the GV type system to provide
the benefits of aliasing with the guarantees of linearity which has been
studied in several systems using a variety of alias control
methods~\cite{Fahndrich:2002,Mazurak:2010:LLT,Tov:2011}. Another approach,
which has been utilised in previous session-based $\lambda$-calculus type
systems, is to support side-effecting functions by allowing functions to close
over channels in the environment~\cite{Gay:2003:STI,Vasconcelos:2006:TCM}. An
interesting extension to the base GV type system would be to incorporate the
closure semantics into the type system which would permit functions to modify
session-typed channels in the environment. The typing rules would be similar
to that described by \citeauthor{Gay:2010:MST} for typing expressions in an
object-oriented setting for session types~\cite{Gay:2010:MST}, e.g. the rule
for lambda abstraction could be:
\begin{gather*}
  \frac{
    \typeds{\Gamma,\typedid{x}{A}}{t}{T}{\Gamma',\typedid{x}{B}}
  }{
    \typeds{\Gamma}{\abs{x}{t}}{\fn{\typestate{A}{B}}{T}}{\Gamma'}
  }\quad
\end{gather*}

The $\Gamma$, $\Gamma '$ environments represent pre and post-states of
variables satisfied by term $t$. The typing for the function argument is
inspired by typestate-oriented programming~\cite{Aldrich:2009} and says that
the function expects a channel behaving according to protocol $A$ and alters
the state of the channel internally producing a channel behaving according to
protocol $B$. It remains to be seen whether such typing rules can be encoded
as described in the GV setting.

These extensions will argue that the framework supports a wide range of
languages and type systems, providing researchers the ability to harness the
power of formal verification in the design, implementation and documentation
of session-based type systems.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background Survey}

%% present an overview of relevant previous work including articles, books,
%% and existing software products. Critically evaluate the strengths and
%% weaknesses of the previous work.

@@

The definition of session types will be important for the remainder:
\input{session-types.tex}

@@

They represent dual intuitionistic linear logic as a sequent calculus
providing logical connectives which have session-based interpretations in
$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL, which have
only two alternatives (cf. multiple alternatives in the standard definition in
\S~\ref{sec:pre}):
\input{sill-branch-choice-rules}

@@

The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
aliased access to linear resources. The type system requires special language
features, e.g. to manage capabilities which can be thought as a form of object
liveness mechanism, so GV typing judgements would need to be extended such
that the capabilities can be included in soundness results. An extension to
the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
would also need to be defined. It is not immediately clear how these features
would assist in the management of session typed channels. The ``focus'' rule
(and linearity generally) enforces an invariant on the focussed object
(channel) such that the type is maintained during the focus operation:
\input{focus-rule}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proposed Approach}\label{sec:approach}

%% state how you propose to solve the software development problem. Show that
%% your proposed approach is feasible, but identify any risks.

I intend to mechanise the calculi described by
\citeauthor{Wadler:2014}~\cite{Wadler:2014} using the Coq proof assistant. In
the spirit of the POPLMARK challenge, I aim to utilise available libraries and
infrastructure as much as
possible. \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a
metatheory library for handling, among other things, association lists
(environments) and free variable representation. I intend to use the version
distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which has been
updated for Coq version 8.4pl4. \citeauthor{Park:2014:MMW} describe a
technique for removing nonlinear contexts from typing judgements. While their
work extended to \fpop it is not clear how to handle an environment containing
both non-linear and linear types as in GV. I wish to maintain as close a
relationship as possible to the paper system presented by
\citeauthor{Wadler:2014}, so separating out the non-linear and linear
components (as in \fpop) is not an option at this stage.

Following \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, I use the
locally nameless first-order representation for binders and free variables,
preventing variable capture issues present in other first-order
representations. Likewise, I make some simplifying assumptions about variable
substitution, namely that \textit{closed} terms (terms with no unbound
indices) are substituted for variables, which avoids having to define lifting
constructs on the substitute term. The typing rules will not contain a rule
for an unbound index so this decision is sound. A consequence is that I must
define a notion of opening language terms involving binders as described
by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
$t$ which is well-typed if $t$ is well-typed when its unbound index,
representing the argument to the abstraction, is replaced by some fresh
variable $x$, written $t^x$. Some terms in GV require multiple binders but I
believe it will be straightforward to extend this notion. Unfortunately, I
must define similar operations on CP terms and handle substitution and opening
on them as well as propositions which can accept propositional variables. I
hope to be able to generalise as much as possible with the help of the
Metatheory library.

Alternatively, I could have chosen a higher-order abstract syntax (HOAS)
approach such as the parametric version (PHOAS) described
by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
on the metalanguage (Coq, in this case) for handling binders and substitution
so has less overhead in managing syntax than the locally nameless
approach. However, variables are now parameters defined at the level of the
metalanguage so can represent terms not valid in the language being
mechanised. Further, a higher-order approach would prevent the use of the
Metatheory library which provides a number of useful tactics for manipulating
typing contexts.

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\cite{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, which
excludes a finite set of variables from being considered as the binder. In
contrast, the traditional ``exists-fresh'' approach, where the binder is only
required to be fresh within the abstraction's body, does not produce strong
enough induction principles in some cases. This change will necessitate
changes to the typing rules of GV and CP. Consider for instance the Connect
rule in GV~\cite{Wadler:2014} (C) and its corresponding cofinite
quantification version (CCO):
\input{with-connect-rule}

The conclusion specifies a variable in (C) but a type in (CCO). Thus, in the
premise for (CCO) one can choose a suitably fresh $x$ in each premise. Note
that, $L$ can be chosen such that $\Psi \cup \Phi \cup \fv(M) \cup \fv(N)
\subseteq L$, where $\fv(M)$ are the set of free variables in term $M$, and
$x$ can be the same in both premises. Similar changes are made to the rules
for CP so that, for example, output which allocates a fresh variable $y$ no
longer names $y$ in the conclusion: \input{cp-output}

Since the name $y$ is assumed to be fresh in the original presentation these
changes do not alter the typing of processes nor the semantics of the
translation from GV to CP. An equivalence between ``exists-fresh'' and
cofinite quantification definitions of the simply-typed $\lambda$-calculus is
given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. In general, I
do not aim to formally prove adequacy of the mechanisation.

While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
may need to alter the approach slightly to avoid tricky formulations. For
instance, instead of providing $n$-ary branch and choice in GV, I provide
only binary versions of these operators to simplify the development. This
change is not restrictive however, since CP itself provides only binary
versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
technical note, I amalgamate all types into one inductive definition and
provide a predicate for restricting typing rules to consider only valid
session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
recursive definitions; session types are types which may contain types as
subcomponents (arguments to send, for example). Unfortunately, handling
mutually inductive definitions in Coq can be quite involved; requiring one to
either rely on the Coq system to provide a stronger mutual induction principle
or defining one manually. Such a definition is possible, but it complicates
elimination via the induction principle.

Another aspect of the encoding is how to define terms. The intrinsic encoding
described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
approach, indexing terms by their type so as to prevent ill-formed terms from
being constructed. However, in the encoding for GV there are issues with using
this encoding: $(1)$ it is not immediately clear how to handle linear contexts
using the de Bruijn variable encoding, since in the intrinsic setting the
environment supports weakening; $(2)$ an intrinsic encoding of terms would be
complicated by the need to enforce that session types occur in certain
instances (by use of a predicate); and $(3)$ well-typed terms require extra
assumptions about binder freshness which cannot reasonably be expressed as a
function type. For these reasons, an intrinsic approach does not offer much
benefit since a well-typed term relation would still need to be
defined. Therefore, I chose an extrinsic encoding for terms and define
well-typed terms as a separate inductive type.

\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
their language proof infrastructure is proportional to the number of binding
constructs. In the case of a simply typed lambda calculus this is not onerous,
but GV has four binding constructs, CP has six and propositions have two. It
may be true that this approach results in too many auxiliary definitions and
operations not relevant to the actual properties one wishes to prove. However,
I believe that once this infrastructure has been defined, the system can be
extended without concern for these binding constructs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Work Plan}

The deliverables can be split into three main areas as indicated in the
following table:

\begin{tabularx}{\linewidth}{|c|X|}\hline
Deliverable \# &  Description \\ \hline
\Deliverable{1}{Mechanisation of Wadler's system}
D1 & Mechanisation of Wadler's system \\ \hline
\Deliverable{2}{Closure side-effecting extension}
\DSubD{2}{1}{Implementation of the GV operational semantics}
\DSubD{2}{2}{Implementation of the side-effecting closure extension}
\Deliverable{3}{Report write-up}
\DSubD{3}{1}{Description of approach, Wadler mechanisation and related work}
\DSubD{3}{2}
      {Description of GV operational semantics (w/o side-effecting closure)}
\DSubD{3}{3}{Description of side-effecting closure and final version}
\end{tabularx}

The current proposal will form part of D3.1 (related work, etc.). The
remaining parts of D3 will be written up as the project progresses, conducted
in tandem with D1 and D2. The estimated work plan is given in the following
table and sub-divides the above deliverable stages into tasks. The tasks
marking the completion of a deliverable are noted. The tasks appear in
chronological order.

\begin{center}
\begin{tabular}{|*{5}{c|}}\hline
\multicolumn{5}{|c|}{\textbf{Estimated Work Plan}}\\ \hline
Task & Start Date & Duration (days) & End Date & Deliverables \\ \hline
\DTask{1}{1}{8th December}{15}{23rd December}{-}
\DTask{3}{1}{20th December}{3}{23rd December}{-}
\DTask{1}{2}{24th December}{17}{10th January}{D1}
\DTask{3}{2}{7th January}{11}{18th January}{D3.1}
\DTask{2}{1}{11th January}{14}{25th January}{-}
\DTask{2}{2}{23th January}{21}{16th Feburary}{D2.1}
\DTask{3}{3}{13th Feburary}{10}{23rd Feburary}{D3.2}
\DTask{2}{3}{17th Feburary}{14}{3rd March}{D2.2}
\DTask{3}{4}{1st March}{14}{15th March}{D3.3}
\DTask{*}{*}{16th March}{39}{24th April}{-}
\end{tabular}
\end{center}

\IDTsk{1}{1} involves completing the GV and CP definitions including reduction
and commuting conversion rules for CP which I aim to complete during
December. This will require definition of infrastructure for handling binders
for both GV, CP channels and CP propositional variables, and inductive
definitions for well-typed terms, commuting conversions and CP principal cut
reductions. So far I have completed some of the definitions for well-typed
terms and CP process rules; the remaining rules require the infrastructure for
substitution and opening of binders. \IDTsk{3}{1} will consist of an initial
structuring and write up of the mechanisation effort of Wadler's system. It
will include a detailed description of the approach taken (which can, in part,
be formed from the current proposal) and the related work (again, partly
formed from the current document). Additionally, it will describe any issues
which have arisen, including changes made to any of the definitions as
presented by Wadler, including the changes already detailed here.

For D1, it remains to define the translation between GV and CP which I
envisage will require several rules on typing derivations; I plan to define
these as inductive datatype constructors but I shall take care to reuse any
suitable libraries that are available, e.g. a library for handling CPS
translations. For this step of the work (\IDTsk{1}{2}), I made a cautious
estimate on duration, accounting some slack time for any delays in preceding
stages of the development. \IDTsk{3}{2} will document the translation phase
and will also be used to document the initial approach for the extension (see
table for overlapping task allocation). \IDTsk{3}{2} represents delivery of
the first draft (D3.1) of the report describing the mechanisation of Wadler's
system but omitting the theoretical treatment for the proposed extension.

For the remainder, I shall look to extend the definitions of GV and CP using
the definitions I have provided. As discussed previously, it would be
interesting to study a system for supporting side-effecting functions (aka
closures) for session channels. In order to achieve this, operational
semantics for GV would need to be defined and the recent draft by
\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
do this~\cite{Lindley:2014:SPS}. \IDTsk{2}{1} is a theoretical investigation
into the operational semantics, deciding whether the system by
\citeauthor{Lindley:2014:SPS} offers a suitable basis for the extension, and
investigating the details of the side-effecting closure extension (assuming an
appropriate base operational semantics for GV). \IDTsk{2}{2} is then the
implementation of the suitable operational semantics (without side-effecting
closure), culminating in the first part of the second deliverable
(D2.1). Following this implementation work, I plan to write up the operational
semantics in the report (\IDTsk{3}{3}) which constitutes the second draft of
the report deliverable (D3.2). Lastly, \IDTsk{2}{3} is an allocation of time
to the task of implementing the extension permitting closures with
side-effects which includes necessary changes to the typing judgements and
translation between GV and CP. In my work plan, I have assumed \IDTsk{2}{2}
implemented an operational semantics that eases this particular extension
which justifies the task allocation time. Both \IDTsk{2}{2} and \IDTsk{2}{3}
may also include revisions to the theoretical treatment started in
\IDTsk{2}{1} as a consequence of initial design oversights. \IDTsk{3}{4}
constitutes the final version of the report, including discussion of the
side-effecting closure extension.

There is a risk of being unable to perform the side-effecting extension as
sketched in \S \ref{sec:approach}. Additionally, there is considerable slack
time identified in the work plan, indicated by \IDTsk{*}{*}. I believe these
task allocation time estimates to be an accurate reflection of the work
involved. However, the feasibility study is necessarily limited in scope, so I
shall review and update the above plan throughout the project as I proceed to
the various milestones I have set out, and if required will alter the above
plans if necessary.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\documentclass{mpaper}

\usepackage{macros}
\usepackage{hyperref}

\usepackage{listings}
\usepackage{lstcoq}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{cmll}
\usepackage{prooftree}
\usepackage{amsmath}
\newcommand{\typestate}[2]{{#1}\gg{#2}}
\usepackage[style=alphabetic,sorting=nymt,natbib=true,
  maxbibnames=100]{biblatex}
\addbibresource{paper.bib}

@@

@@

@@

@@

\title{Mechanisation of Session-based Lambda Calculus Type Systems}
\author{Craig McLaughlin}
\matricnum{1002524}

\maketitle

\begin{abstract}
According to Simon Peyton Jones, an abstract should address
four key questions. First, what is the problem that this
paper tackles? Second, why is this an interesting problem?
Third, what is the solution this paper proposes?
Finally, why is the proposed solution a good one?
\end{abstract}

\section{Introduction}\label{intro}

@@

\section{Background}

@@

The definition of session types will be important for the remainder:
\input{../proposal/session-types.tex}

@@

They represent dual intuitionistic linear logic as a sequent calculus
providing logical connectives which have session-based interpretations in
$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL, which have
only two alternatives (cf. multiple alternatives in the standard definition in
\S~\ref{sec:pre}):
\input{../proposal/sill-branch-choice-rules}

@@

The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
aliased access to linear resources. The type system requires special language
features, e.g. to manage capabilities which can be thought as a form of object
liveness mechanism, so GV typing judgements would need to be extended such
that the capabilities can be included in soundness results. An extension to
the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
would also need to be defined. It is not immediately clear how these features
would assist in the management of session typed channels. The ``focus'' rule
(and linearity generally) enforces an invariant on the focussed object
(channel) such that the type is maintained during the focus operation:
\input{../proposal/focus-rule}

@@

\section{Evaluation}

Graphs are always good. I recommend getting to grips with Matlab, R or
gnuplot rather than exporting horribly Excel bitmapped graphs.

\begin{quotation}
 The Assyrian came down like the wolf on the fold,
 And his cohorts were gleaming in purple and gold;
 And the sheen of their spears was like stars on the sea,
 When the blue wave rolls nightly on deep Galilee.

 Like the leaves of the forest when Summer is green,
 That host with their banners at sunset were seen:
 Like the leaves of the forest when Autumn hath blown,
 That host on the morrow lay withered and strown.
\end{quotation}

\section{Conclusions}

{\bf Acknowledgments.}
This is optional; it is a location for you to thank people
